\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-20 doc/incertae_sedis.tex

\author{Benjamin D. Redelings and Mark T. Holder}

\maketitle

@@

The supertree method described by \citet{redelings2017supertree} is
able to use a set of phylogenetic estimates and taxonomy to produce a
supertree with clades names according to the taxonomy. However, the
method was not able to accommodate the fact that taxonomies frequently
contain nodes with uncertain placement. These nodes are often labelled
``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
The common interpretation of such taxa is that they may not be moved
further towards the root, but may be moved into their sibling taxa.
For example, a genus with a sibling that is a family may be annotated
as ``\emph{incertae familia}'', indicating that it is unclear which
family the genus should be placed in. \emph{Incertae sedis} taxa
frequently occur when a specimen is identified down to a given
taxonomic level, but no further. Extinct taxa are often \emph{incertae
sedis}. OTT is constructed from several source taxonomies, and these
sources include various ways of indicating that the position of a
taxon within the hierarchy is uncertain. OTT uses a series of
``flags'' to annotate these taxa. The supertree of by
\citet{redelings2017supertree} simply pruned these taxa from the
taxonomy and input trees. Thus, the final supertree lacked taxa which
were \emph{incertae sedis} or the descendants of such taxa.

@@

The taxonomy tree, $\taxonomy_{p}$, is derived from OTT. As mentioned
above, OTT not only contains a taxonomic hierarchy, but also contains
taxonomic labels and annotations (referred to as ``flags'') on the
taxa to carry extra information. The entire hiearchy encoded by OTT is
a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
labels $\mathcal{N}$ that are called taxon names. We will decompose
$\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
}and higher taxa $\mathcal{H}$. We will use $\mathcal{U}$ to refer to
the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
annotated with the \emph{incertae sedis }property. Other annotations
identify dubious taxa and taxonomic names that are unwanted artifacts
of the process of building OTT. Thus, the taxonomic tree which is used
as an input to the supertree algorithm is a pruned version of
$\taxonomy^{\ast}$. For our purposes, the relevant information from
the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis
}taxa. In previous work, taxa in $\mathcal{U}$ and all of their descendants were all pruned from
the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT. In
the present work, we consider supertree approaches that operate on a
taxonomic tree $\taxonomy$, which is produced from
$\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.

@@

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we must arrive at an operational definition of the meaning
of the \emph{incertae sedis} label. The core meaning of the phrase is
that the annotated taxon may actually be a member of one of the taxa
that appear as siblings to it in the heirarchical representation of
the taxonomy. In the most expansive interpretation, the author of the
taxonomy is stating that the \emph{incertae sedis} taxon could be
placed anywhere inside one of the sibling taxa or their descendants.
However, frequently only a subset of the possible placement points
would be viewed as plausible. Ideally, the \emph{incertae sedis}
annotation for some taxon \emph{x} would be accompanied by a specific
statement about which other taxa might include \emph{x}. Detailed
readings of the character argumentation used by taxonomists could
provide this sort of specific information in many cases, but we lack
such information on the scale of the entire OTT. Here we consider a
few operational approaches for handling the \emph{incertae sedis}
label.

@@

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy. This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families. The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree. For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree (Fig. \ref{fig:Handling-incertae-sedis}d),
but not the taxonomy (Fig. \ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$. Furthermore, the synthetic tree additionally places
$C$ within $AB$.

@@

The addition of \emph{incertae sedis} taxa allows new types of
conflict between input trees. For example, different input trees might
place an incertae sedis taxon in conflicting locations. This is
illustrated in Figure \ref{fig:placement-example1}, where the IS
taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
$\tau_{1}$ and as sister ott3 by phylogeny $\tau_{2}$.

>>>>

\author{Benjamin D.~Redelings and Mark T.~Holder} 
\maketitle

@@

The supertree method described by \citet{redelings2017supertree} is

able to use a set of phylogenetic estimates and taxonomy to produce a
supertree with clades names according to the taxonomy.
However, the
method was not able to accommodate the fact that taxonomies frequently
contain nodes with uncertain placement.
These nodes are often labelled
``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
The common interpretation of such taxa is that they may not be moved
further towards the root, but may be moved into their sibling taxa.
For example, a genus with a sibling that is a family may be annotated
as ``\emph{incertae familia}'', indicating that it is unclear which
family the genus should be placed in.
\emph{Incertae sedis} taxa
frequently occur when a specimen is identified down to a given
taxonomic level, but no further.
Extinct taxa are often \emph{incertae
sedis}.
OTT is constructed from several source taxonomies, and these
sources include various ways of indicating that the position of a
taxon within the hierarchy is uncertain.
OTT uses a series of
``flags'' to annotate these taxa.
The supertree of by
\citet{redelings2017supertree} simply pruned these taxa from the
taxonomy and input trees.
Thus, the final supertree lacked taxa which
were \emph{incertae sedis} or the descendants of such taxa.

@@

The taxonomy tree, $\taxonomy_{p}$, is derived from OTT.
As mentioned
above, OTT not only contains a taxonomic hierarchy, but also contains
taxonomic labels and annotations (referred to as ``flags'') on the
taxa to carry extra information.
The entire hiearchy encoded by OTT is
a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
labels $\mathcal{N}$ that are called taxon names.
We will decompose
$\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
}and higher taxa $\mathcal{H}$.
We will use $\mathcal{U}$ to refer to
the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
annotated with the \emph{incertae sedis }property.
Other annotations
identify dubious taxa and taxonomic names that are unwanted artifacts
of the process of building OTT.
Thus, the taxonomic tree which is used
as an input to the supertree algorithm is pruned version of
$\taxonomy^{\ast}$.
For our purposes, the relevant information from
the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis} taxa.
In previous work, taxa in $\mathcal{U}$ and all of their descendants 
were all pruned from
the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT.
In
the present work, we consider supertree approaches that operate on a
taxonomic tree $\taxonomy$, which is produced from
$\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.

@@

In order to incorporate \emph{incertae sedis }taxa into a supertree
analyses, we must arrive at an operational definition of the meaning
of the \emph{incertae sedis} label.
The core meaning of the phrase is
that the annotated taxon may actually be a member of one of the taxa
that appear as siblings to it in the heirarchical representation of
the taxonomy.
In the most expansive interpretation, the author of the
taxonomy is stating that the \emph{incertae sedis} taxon could be
placed anywhere inside one of the sibling taxa or their descendants.
However, frequently only a subset of the possible placement points
would be viewed as plausible.
Ideally, the \emph{incertae sedis}
annotation some taxon $x$ would be accompanied by a specific
statement about which other taxa might include $x$.
Detailed
readings of the character argumentation used by taxonomists could
provide this sort of specific information in many cases, but we lack
such information on the scale of the entire OTT.
Here we consider a
few operational approaches for handling the \emph{incertae sedis}
label.

@@

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy.
This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families.
The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree.
For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree 
(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
(Fig.\ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$. Furthermore, the synthetic tree additionally places
$C$ within $AB$.

@@

The addition of \emph{incertae sedis} taxa allows new types of
conflict between input trees.
For example, different input trees might
place an incertae sedis taxon in conflicting locations.
This is
illustrated in Figure \ref{fig :placement-example1}, where the IS
taxon (ott7,ott6)ott10 is placed as sister to ott1 by phylogeny
$\tau_{1}$ and as sister ott3 by phylogeny $\tau_{2}$.
